---
title: "CodeBook"
author: "Neeket Agarwal"
date: "4/28/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Human Activity Recognition Using Smartphones Dataset

The experiments have been carried out with a group of 30 volunteers within an age bracket of 19-48 years. Each person performed six activities (WALKING, WALKING_UPSTAIRS, WALKING_DOWNSTAIRS, SITTING, STANDING, LAYING) wearing a smartphone (Samsung Galaxy S II) on the waist. Using its embedded accelerometer and gyroscope, we captured 3-axial linear acceleration and 3-axial angular velocity at a constant rate of 50Hz. The experiments have been video-recorded to label the data manually. The obtained dataset has been randomly partitioned into two sets, where 70% of the volunteers was selected for generating the training data and 30% the test data. 

The sensor signals (accelerometer and gyroscope) were pre-processed by applying noise filters and then sampled in fixed-width sliding windows of 2.56 sec and 50% overlap (128 readings/window). The sensor acceleration signal, which has gravitational and body motion components, was separated using a Butterworth low-pass filter into body acceleration and gravity. The gravitational force is assumed to have only low frequency components, therefore a filter with 0.3 Hz cutoff frequency was used. From each window, a vector of features was obtained by calculating variables from the time and frequency domain.

For each record the datasets provide:
- Triaxial acceleration from the accelerometer (total acceleration) and the estimated body acceleration.
- Triaxial Angular velocity from the gyroscope. 
- A 561-feature vector with time and frequency domain variables. 
- Its activity label. 
- An identifier of the subject who carried out the experiment.

## Variables in the tidy datasets

The tidied data set (subdata) contains 75 variables for the each 30 subject IDs and 6 activities. 
```{r Number of variables}
length(subdata)
```
```{r Number of unique Subject ID}
length(unique(subdata$Subject.ID))
```
```{r Number of unique activities}
length(unique(subdata$activity))
```

The following table describes the snippets which forms the variables and its calculations.

Variable/Snippet |  Description 
------------- | ----------------------------------------
Subject.ID | Subject ID for each of the 30 measurement. 
activity | One of the 6 Activity performed during the measurement, i.e., walking, walking upstairs, walking downstairs, sitting, standing, and laying.  
t | Time domain signals collected at a rate of 50 Hz. Data were filtered using a median filter and a 3rd order low pass Butterworth filter with a corner frequency of 20 Hz to remove noise.
f | Frequency domain signals calculated by applying a Fast Fourier Transform (FFT) to the time domain signal.
Body | The portion of the accelerometer measurement attributed to the subject.
Gravity | The portion of the accelerometer measurement attributed to gravity.
Acc | A linear acceleration measurement taken by the accelerometer. 
Gyro | An angular acceleration measurement taken by gyroscope
Jerk | Jerk was calculated by taking the derivative in time of the measurement.
Mag | The magnitude of the three-dimensional signals were calculated using the Euclidean norm.
Mean | The mean of a given measurement.
Std | The standard deviation of a given measurement.
X, Y, or Z | Indicates the three-dimensional direction of the measurement.
Angle | Angular value between vectors. 

The full variable list can be obtained by combining these various snippets.

## List of variables

```{r List of variables}
names(subdata)
```
## Summaries calculated
Data from various files are read into the following data frames:
- data: Contains data from test and train x-measurements 
- label: Contains activity labels for test and train measurements 
- subject: Contains Subject ID for test and train measurements
- activity.name : Contains activity ID to name mapping 
- master.data: Contains all measurement, Subject ID and activity data 
- subdata: Contains only the measurements on the mean and standard deviation for each measurement 
- avg.data: Contains independent tidy data set with the average of each variable for each activity and each subject. 
- subdata.txt: Contains subdata in txt format in Output_data folder
- avg_data.txt: Contains avg.data in txt format in Output_data folder


